On the specific recursive structure of iterated parametricity

### 2025-11-09 paper/paper.tex
paper: introduce fulldeps to include E, fix errors

\newcommandx{\deps}[2][1,2]{\prim{deps}[][#1][#2][][]}
\newcommandx{\depscohs}[2][1,2]{\prim{depscohs}[][#1][#2][][]}
\newcommandx{\depscoht}[2][1,2]{\prim{depscoht}[][#1][#2][][]}
\newcommandx{\depstype}[2][1,2]{\prim{depstype}[][#1][#2][][]}
\newcommandx{\depscohstype}[2][1,2]{\prim{depscohstype}[][#1][#2][][]}
\newcommandx{\depscohttype}[2][1,2]{\prim{depscohttype}[][#1][#2][][]}

@@

At this stage, it is convenient to use an abbreviation to combine sequences $\framep[n][0,..,p-1]$, $\painting[n][0,...,p-1]$ and $\restrf[n][0,...,p-1]$ in one block, which we call $\deps[n][0,...p-1]$:

\begin{align*}
  \deps[n][0,..,p-1] \defeq
  \left(\begin{array}{l}
            \framep[n][0,...,p-1] :\frametype[n][0,...,p-1] \\
            \painting[n][0,...,p-1] :\paintingtype[n][0,...,p-1]
            (\framep[n][0,...,p-1])                         \\
            \restrf[n][0,...,p-1]  :\restrftype[n][0,...,p-1]
            \left(\begin{array}{l}
                \framep[n][0,...,p-1]   \\
                \painting[n][0,...,p-1] \\
              \end{array}\right)                   \\
          \end{array}\right)
\end{align*}

so that the type of $\framep[n][0,...,p]$ can be shortened into $\depstype[n][0,..,p-1] \rightarrow \frametype[n][0,...,p]$ and similarly for $\layertype[n][p]$ and $\layer[n][p]$.
Later, we shall write $\deps.\framep$, $\deps.\painting$, $\deps.\restrf$ for the corresponding projections of a $\deps$. By abuse of notation, we also write $\unitpoint$ for the only inhabitant of $\depstype[n][0,..,0-1]$.

@@

We can now formally define $\painting[n][p]$ by recursion on $n-p$:
\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \painting[n][p]          &
    \multicolumn{3}{l}{
      \left(\begin{array}{l}
                \deps[n-1][0,...,n-1]:\depstype[n-1][0,...,n-1] \\
                E:\framep[n-1][n-1]\rightarrow \U               \\
              \end{array}\right) : \framep[n-1][n-1]\rightarrow \U
    }                                                            \\
    \painting[n][n]          &
    \left(\begin{array}{l}
              \deps[n-1][0,...,n-1] \\
              E                     \\
            \end{array}\right) & \defeq & E                        \\
    \painting[n][p](p<n)     &
    \left(\begin{array}{l}
              \deps[n-1][0,...,n-1] \\
              E                     \\
            \end{array}\right) & \defeq & d \mapsto
    \left(\begin{array}{l}
              \Sigma l:\layer(\deps[n-1][0,...,p-1]). \\
              \painting[n][p+1,...,n-1]
              \left(\begin{array}{l}
                  \deps[n-1][0,...,n-1] \\
                  E                     \\
                \end{array}\right)(d,l)           \\
            \end{array}\right) \\
  \end{array}
\end{equation*}

@@

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \painting[n][0,...,p]                              &
    \left(\begin{array}{l}
              \deps[n-1][0,...,n-1]:\depstype[n-1][0,...,n-1] \\
              E:\framep[n-1][n-1]\rightarrow \U               \\
            \end{array}\right) & :      & \paintingtype[n][0,...,p](\framep[n][0,...,p](\deps[n-1][0,...,p])) \\
    \painting[n][0,...,0-1]                            &
    \left(\begin{array}{l}
              \unitpoint \\
              E          \\
            \end{array}\right)                             & \defeq & \unittype                               \\
    \painting[n][0,...,p]                              &
    \begin{array}{l}
      \deps[n-1][0,...,p-1] \\
      E                     \\
    \end{array}                           & \defeq &
    \left(\painting[n][0,...,p-1]
    \left(\begin{array}{l}
                \deps[n-1][0,...,n-1] \\
                E                     \\
              \end{array}\right),\,
    \painting[n][p]
    \left(\begin{array}{l}
                \deps[n-1][0,...,n-1] \\
                E                     \\
              \end{array}\right)\right)                                                                         \\
  \end{array}
\end{equation*}

@@

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \restrptype[n][0,...,p]                    & (\deps[n][0,...,n]:\depstype[n][0,...,n]) & :      & \U                                        \\
    \restrptype[n][]                           & \unitpoint                                & \defeq & \unittype                                 \\
    \restrptype[n][0,...,p](\deps[n][0,...,n]) & (\deps[n][0,...,n])                       & \defeq & \restrptype[n+1][0,...,p-1] \times \left(
    \begin{array}{l}
        \Pi q\leq n.\Pi \epsilon.\Pi d:\framep(\deps[n][0,...,p]). \\
        \painting[n][p](\deps[n][0,...,n])(d) \rightarrow          \\ \deps.\painting[n][p](\deps.\restrf[n][p][q][\epsilon](d))\end{array}
    \right)                                                                                                                                     \\
  \end{array}
\end{equation*}

@@

                \deps[n][0,...,n]:\depstype[n][0,...,n] \\

@@

@@

                \deps.\framep[n][0,...,p]   \\
                \deps.\painting[n][0,...,p] \\

@@

@@

                  \deps[n+1][0,...,n]     \\

@@

              \deps[n][0,...,n]:\depstype[n][0,...,n]                              \\

@@

              \deps[n][0,...,n]:\depstype[n][0,...,n]                              \\

@@

            \framep[n][p](\deps[n][0,...,p])   \\
            \painting[n][p](\deps[n][0,...,n]) \\

@@

                      \deps[n][0,...,p]     \\

@@

        \deps.\layer[n][p](\deps[n][0,...,n])(\restrf[n][p]

@@

              \deps[n][0,...,n]:\depstype[n][0,...,n]                              \\
              \cohf[n][0,...,p-1]:\cohftype[n][0,...,p-1]                          \\

@@

\begin{align*}
  \depscohstype[n][0,..,p-1] \defeq
  \left(\begin{array}{l}
            \deps[n][0,...,n] :\depstype[n][0,...,n]        \\
            \restrp[n][0,...p-1]: \restrptype[n][0,...,p-1] \\
            \cohf[n][0,...,p-1]: \cohftype[n][0,...,p-1]    \\
          \end{array}\right)
\end{align*}
so that the type of $\restrf[n][0,...,p]$ can be rephrased as $\depscohstype[n][0,..,p-1] \rightarrow \restrftype[n][0,...,p]      \left(\begin{array}{l}
      \deps.\framep[n][0,...,p]   \\
      \deps.\painting[n][0,...,p] \\
    \end{array}\right)$, and similarly for $\restrltype$ and $\restrl$. Later, we shall write $\depscohs.\deps$, $\depscohs.\restrp$, and $\depscohs.\cohf$ for the corresponding projections of an element $\depscohs$ in $\depscohstype$. By abuse of notation, we also write $\unitpoint$ for the only inhabitant of $\depscohstype[n][0,..,0-1]$.

@@

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \cohptype[n][0,...,p] & (\depscohs[n][0,...,n]:\depscohstype[n][0,...,n]) & :      & \U                                      \\
    \cohptype[n][]        & \unitpoint                                        & \defeq & \unittype                               \\
    \cohptype[n][0,...,p] & (\depscohs[n][0,...,n])                           & \defeq & \cohptype[n+1][0,...,p-1] \times \left(
    \begin{array}{l}
        \Pi q\leq n.\Pi r\leq q.\Pi \epsilon.\Pi \omega.\Pi d:\framep(\depscohs[n][0,...,p]). \\
        \painting[n][p](\depscohs[n][0,...,n])(d) \rightarrow                                 \\ \depscohs.\painting[n][p](\depscohs.\restrf[n][p][q][\epsilon](d))\end{array}
    \right)                                                                                                                      \\
  \end{array}
\end{equation*}

@@

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \cohp[n][p]                                        &
    \left(\begin{array}{l}
              \depscoht[n][0,...,n]:\depscohttype[n][0,...,n] \\
              E:\framep[n-2][n-2]\rightarrow \U               \\
              E':\framep[n-1][n-1]\rightarrow \U              \\
              E'':\framep[n][n]\rightarrow \U                 \\
            \end{array}\right) & :      &
    \left(
    \begin{array}{l}
        \Pi q\leq n. \Pi \epsilon. \Pi d:\framep[n][p]
        \left(\begin{array}{l}
                  \overline{\depscohs}(\depscoht) \\
                  E':\framep[n][n]\rightarrow \U  \\
                \end{array}\right).   \\
        \painting[n][p]
        \left(\begin{array}{l}
                  \overline{\depscohs}(\depscoht) \\
                  E':\framep[n][n]\rightarrow \U  \\
                \end{array}\right)(d) \rightarrow \\
        \painting[n-1][p]
        \left(\begin{array}{l}
                  \depscoht[n][0,...,n].\deps       \\
                  E:\framep[n-1][n-1]\rightarrow \U \\
                \end{array}\right)(\cohf[n][p](\depscoht[n][0,...,n].deps)(d))\end{array}\right)                                                  \\
    \cohp[n][n]                                        &
    \left(\begin{array}{l}
              \depscoht[n][0,...,n] \\
              E                     \\
              E'                    \\
            \end{array}\right)                           & \defeq & \lambda q\leq n.\lambda \epsilon. \lambda d. \lambda (l,c). l\epsilon \\
    \cohp[n][p](p<n)                                   &
    \left(\begin{array}{l}
              \depscoht[n][0,...,n] \\
              E                     \\
              E'                    \\
            \end{array}\right)                           & \defeq & \lambda q\leq n.\lambda \epsilon. \lambda d. \lambda (l,c).
    \left(\begin{array}{l}
              \cohl[n][p][q][\epsilon](\depscoht[n-1][0,...,p-1])(l),            \\
              \cohp[n][p][q][\epsilon]
              \left(\begin{array}{l}
                  \depscoht[n][0,...,n] \\
                  E                     \\
                  E'                    \\
                \end{array}\right)(\cohf[n][p](\depscoht[n][0,...,p])(d))(c) \\
            \end{array}\right)                                                           \\
  \end{array}
\end{equation*}

>>>>

\newcommandx{\deps}[2][1,2]{\prim{deps}[][#1][#2][][]}
\newcommandx{\fulldeps}[2][1,2]{\prim{fulldeps}[][#1][][][]}
\newcommandx{\depscohs}[2][1,2]{\prim{depscohs}[][#1][#2][][]}
\newcommandx{\depscoht}[2][1,2]{\prim{depscoht}[][#1][#2][][]}
\newcommandx{\depstype}[2][1,2]{\prim{depstype}[][#1][#2][][]}
\newcommandx{\fulldepstype}[2][1,2]{\prim{fulldepstype}[][#1][][][]}
\newcommandx{\depscohstype}[2][1,2]{\prim{depscohstype}[][#1][#2][][]}
\newcommandx{\depscohttype}[2][1,2]{\prim{depscohttype}[][#1][#2][][]}

@@

At this stage, it is convenient to use an abbreviation to combine sequences $\framep[n][0,..,p-1]$, $\painting[n][0,...,p-1]$ and $\restrf[n][0,...,p-1]$ in one block, which we call $\depstype[n][0,...p-1]$:

\begin{align*}
  \depstype[n][0,..,p-1] \defeq
  \left(\begin{array}{l}
            \framep[n][0,...,p-1] :\frametype[n][0,...,p-1] \\
            \painting[n][0,...,p-1] :\paintingtype[n][0,...,p-1]
            (\framep[n][0,...,p-1])                         \\
            \restrf[n][0,...,p-1]  :\restrftype[n][0,...,p-1]
            \left(\begin{array}{l}
                \framep[n][0,...,p-1]   \\
                \painting[n][0,...,p-1] \\
              \end{array}\right)                   \\
          \end{array}\right)
\end{align*}
so that the type of $\framep[n][0,...,p]$ can be shortened into $\depstype[n][0,..,p-1] \rightarrow \frametype[n][0,...,p]$ and similarly for $\layertype[n][p]$ and $\layer[n][p]$.
Later, we shall write $\deps.\framep$, $\deps.\painting$, $\deps.\restrf$ for the corresponding projections of a $\deps$. By abuse of notation, we also write $\unitpoint$ for the only inhabitant of $\depstype[n][0,..,0-1]$.

@@

Let us introduce another abbreviation to combine a full $\deps[n][0,n]$ with a family $E:\framep[n-1][n-1]\rightarrow \U$:
\begin{align*}
  \fulldepstype[n] \defeq
  \left(\begin{array}{l}
            \deps[n][0,...,n-1] :\depstype[n][0,...,n-1] \\
            E:\frametype[n-1][n-1]\rightarrow \U         \\
          \end{array}\right)
\end{align*}
TODO: say the convention on projection in the Introduction

Also, when $n$ is 0, we write $E$ to denote an element $(\unitpoint, E)$ of $\fulldepstype[0]$.

We can now formally define $\painting[n][p]$ by recursion on $n-p$:
\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \painting[n][p]      &
    (\fulldeps[n-1]:\fulldepstype[n-1])           \\
    \painting[n][n]      &
    (\fulldeps[n-1])     & \defeq & \fulldeps.E   \\
    \painting[n][p](p<n) &
    (\fulldeps[n-1])     & \defeq & d \mapsto
    \left(\begin{array}{l}
              \Sigma l:\layer(\deps[n-1][0,...,p-1]). \\
              \painting[n][p+1,...,n-1]
              (\fulldeps[n-1])(d,l)                   \\
            \end{array}\right) \\
  \end{array}
\end{equation*}

@@

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \painting[n][0,...,p]   &
    (\fulldeps[n-1])        & :      & \paintingtype[n][0,...,p](\framep[n][0,...,p](\fulldeps.\deps[n-1][0,...,p])) \\
    \painting[n][0,...,0-1] &
    \left(\begin{array}{l}
              \unitpoint \\
              E          \\
            \end{array}\right)  & \defeq & \unittype                                                                   \\
    \painting[n][0,...,p]   &
    (\fulldeps[n-1])                                                                                                 \\
                            & \defeq &
    \left(\painting[n][0,...,p-1]
      (\fulldeps[n-1]),\,
    \painting[n][p]
    (\fulldeps[n-1])\right)                                                                                          \\
  \end{array}
\end{equation*}

@@

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \restrptype[n][0,...,p]             &
    (\fulldeps[n-1]:\fulldepstype[n-1]) & :          & \U                                                    \\
    \restrptype[n][]                    & \unitpoint & \defeq                                    & \unittype \\
    \restrptype[n][0,...,p]             &
    (\fulldeps[n-1])
                                        & \defeq     & \restrptype[n+1][0,...,p-1] \times \left(
    \begin{array}{l}
        \Pi q\leq n.\Pi \epsilon.\Pi d:\framep(\deps[n][0,...,p]). \\
        \painting[n][p]
        (\fulldeps[n-1])(d) \rightarrow                            \\ \fulldeps.\deps.\painting[n][p](\fulldeps.\deps.\restrf[n][p][q][\epsilon](d))\end{array}
    \right)                                                                                                  \\
  \end{array}
\end{equation*}

@@

                \fulldeps[n]:\depstype[n]   \\

@@

                    \fulldeps[n] \\

@@

                \fulldeps.\deps.\framep[n][0,...,p]   \\
                \fulldeps.\deps.\painting[n][0,...,p] \\

@@

              E          \\

@@

                  \fulldeps[n+1]          \\

@@

              \fulldeps[n]:\fulldepstype[n]                                 \\
              \restrp[n][0,...,p-1]:\restrptype[n][0,...,p-1](\fulldeps[n]) \\

@@

              \fulldeps[n]:\fulldepstype[n]                                        \\

@@

            \framep[n][p](\fulldeps.\deps[n][0,...,p])   \\
            \painting[n][p](\fulldeps.\deps[n][0,...,n]) \\

@@

                      \fulldeps.\deps[n][0,...,p] \\

@@

        \fulldeps.\deps.\layer[n][p](\deps[n][0,...,n])(\restrf[n][p]

@@

              \fulldeps[n]:\fulldepstype[n]                                 \\
              \restrp[n][0,...,p-1]:\restrptype[n][0,...,p-1](\fulldeps[n]) \\
              \cohf[n][0,...,p-1]:\cohftype[n][0,...,p-1]
                  \fulldeps[n]          \\

@@

\begin{align*}
  \depscohstype[n][0,..,p-1] \defeq
  \left(\begin{array}{l}
            \deps[n][0,...,n] :\depstype[n][0,...,n]        \\
            \restrp[n][0,...p-1]: \restrptype[n][0,...,p-1] \\
            \cohf[n][0,...,p-1]: \cohftype[n][0,...,p-1]    \\
          \end{array}\right)
\end{align*}
so that the type of $\restrf[n][0,...,p]$ can be rephrased as $\Pi \depscohs[n][0,...,p-1]:\depscohstype[n][0,..,p-1]. \restrftype[n][0,...,p]      \left(\begin{array}{l}
      \depscohs.\deps.\framep[n][0,...,p]   \\
      \depscohs.\deps.\painting[n][0,...,p] \\
    \end{array}\right)$, and similarly for $\restrltype$ and $\restrl$. Later, we shall write $\depscohs.\deps$, $\depscohs.\restrp$, and $\depscohs.\cohf$ for the corresponding projections of an element $\depscohs$ in $\depscohstype$. By abuse of notation, we also write $\unitpoint$ for the only inhabitant of $\depscohstype[n][0,..,0-1]$.

@@

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \cohptype[n][0,...,p] & (\depscohs[n][0,...,n]:\depscohstype[n][0,...,n]) & :      & \U                                      \\
    \cohptype[n][]        & \unitpoint                                        & \defeq & \unittype                               \\
    \cohptype[n][0,...,p] & (\depscohs[n][0,...,n])                           & \defeq & \cohptype[n+1][0,...,p-1] \times \left(
    \begin{array}{l}
        \Pi q\leq n.\,\Pi r\leq q.\,\Pi \epsilon.\,\Pi \omega. \\
        \depscohs.\restrp[n][p][q][\epsilon](\restrp[n+1][p][r][\omega]
        \left(\begin{array}{l}
                  \depscohs[n+1][0,...,n] \\
                  \cohp[n+1][0,...,p-1]   \\
                \end{array}\right)(d)) =                         \\
        \depscohs.\restrp[n][p][r][\omega](\restrp[n+1][p][q+1][\epsilon]
        \left(\begin{array}{l}
                  \depscohs[n+1][0,...,n] \\
                  \cohp[n+1][0,...,p-1]   \\
                \end{array}\right)(d))
      \end{array}
    \right)                                                                                                                      \\
  \end{array}
\end{equation*}

@@

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \cohp[n][p]                                        &
    \left(\begin{array}{l}
              \depscoht[n][0,...,n]:\depscohttype[n][0,...,n] \\
              E:\framep[n-2][n-2]\rightarrow \U               \\
              E':\framep[n-1][n-1]\rightarrow \U              \\
              E'':\framep[n][n]\rightarrow \U                 \\
            \end{array}\right) & :      &
    \left(
    \begin{array}{l}
        \Pi q\leq n.\,\Pi r\leq q.\,\Pi \epsilon.\,\Pi \omega. \\
        \depscohs.\restrp[n][p][q][\epsilon](\restrp[n+1][p][r][\omega]
        \left(\begin{array}{l}
                  \depscohs[n+1][0,...,n] \\
                  \cohp[n+1][0,...,p-1]   \\
                \end{array}\right)(d)) =                         \\
        \depscohs.\restrp[n][p][r][\omega](\restrp[n+1][p][q+1][\epsilon]
        \left(\begin{array}{l}
                  \depscohs[n+1][0,...,n] \\
                  \cohp[n+1][0,...,p-1]   \\
                \end{array}\right)(d))
      \end{array}\right)                                                                                                       \\
    \cohp[n][n]                                        &
    \left(\begin{array}{l}
              \depscoht[n][0,...,n] \\
              E                     \\
              E'                    \\
            \end{array}\right)                           & \defeq & \lambda q\leq n.\lambda r\leq q.\lambda \epsilon.\lambda \omega.\lambda d. \lambda (l,c). l\epsilon \\
    \cohp[n][p](p<n)                                   &
    \left(\begin{array}{l}
              \depscoht[n][0,...,n] \\
              E                     \\
              E'                    \\
            \end{array}\right)                           & \defeq & \lambda q\leq n.\lambda r\leq q.\lambda \epsilon.\lambda \omega.\lambda d. \lambda (l,c).
    \left(\begin{array}{l}
              \cohl[n][p][q][\epsilon](\depscoht[n-1][0,...,p-1])(l),            \\
              \cohp[n][p][q][\epsilon]
              \left(\begin{array}{l}
                  \depscoht[n][0,...,n] \\
                  E                     \\
                  E'                    \\
                \end{array}\right)(\cohf[n][p](\depscoht[n][0,...,p])(d))(c) \\
            \end{array}\right)                                                                                         \\
  \end{array}
\end{equation*}
